---
title: 'A Regionalized SUNY Impact Assessment'
author: "Moira O'Neill"
date: "12/5/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The State University of New York (SUNY) is one of the largest public higher education systems in the country, with 64 total institutions spread across 10 state regions.  These regions are highly varied, and include a megacity, the rural periphery, and several old industrial cities experiencing population loss.  The institutions are varied too, and include research universities, comprehensive colleges, community colleges, and technical institutes.

SUNY's heterogeneous geographic and institutional structure is designed to facilitate access and cultivate human capital at several degree levels.  This is explicit in SUNY's stated mission: to recognize the "fundamental role of its responsibilities in undergraduate education and [to provide] a full range of graduate and professional education that reflects the opportunity for individual choice and the needs of society" (suny.edu).

How well does the SUNY system fulfill this mission at the regional and state scales?  The following study offers a preliminary assessment of SUNY's role in attracting, forming, and retaining human capital in different parts of the state.  The purpose of this exercise is 1) to identify strengths and weaknesses in the system, and 2) to generate research questions for further inquiry.

## Methods and Materials

This project uses data compiled from the Census Tigris package, US Census American Community survey, the Rockefeller Center, and the SUNY gradwages dashboard.  To explore the various roles SUNY plays in each region, I use a series of histograms, line, plot, and bar graphs, some with reference lines, to visually convey the different roles the SUNY system plays within each region and within the state.

```{r packages, message = F, results = "hide", echo = TRUE}

#load necessary packages

library(readr)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(RColorBrewer)
library(tidycensus)
library(readr)
library(knitr)
library(kableExtra)
library(latex2exp)
library(egg)
library(gridExtra)
library(lattice)
library(spData)
library(sf)
library(tidyverse)
library(tigris)
library(rgdal)
library(sp)
library(rgeos)
library(ggmap)
library(plotly)

census_api_key("7912510ff91a51de560cf1e9160860875f062880")

```


## Part I.  Who SUNY serves

SUNY has campuses in every part of the state.  Institutions tend to congregate near one another in the state's largest metropolitan areas: New York City and its environs, the capital region of Albany, Syracuse in Central New York, and Buffalo in Western New York.  In each of these areas, comprehensive college and community or technical colleges complement one or more doctoral universities.  


```{r map, message = F, results = "hide", echo = TRUE}

#load new york counties from tigris
ny <- counties("NY")
ny <- counties("NY", cb = TRUE, resolution = "5m")

#load regions and school locations
regions <- read_csv("Census_geography.csv")

#join region attribute with spatial data
ny_regions <- merge(ny, regions, by.x = "NAME", by.y = "County")
class(ny_regions)

#convert sp to sf
ny_regions <- st_as_sf(ny_regions)

#dissolve county borders
ny_regions_dissolved <-
    ny_regions %>%
    group_by(Region) %>%
  summarise(status = 'dissolved')

#load school locations
schools <- read_csv("SUNY_locations_and_student_counts.csv")
#join school locations to ny regions
schools2 <- merge(ny_regions_dissolved, schools, by = "Region")

#plot regions on map
ggplot() +
  geom_sf(ny_regions_dissolved, mapping = aes(fill = Region))+
  scale_fill_brewer(palette = "Paired")+
  geom_point(data = schools, mapping = aes(x= lon, y = lat, shape = Level))+
  ggtitle("SUNY Institutional Distribution by Region")+
  theme_bw()+
  theme(axis.title.x = element_blank(),
           axis.title.y = element_blank()
  )

  
```

The stacked bar plot below demonstrates that a majority of SUNY students attend college within their own region.  This is likely due to cost and information symmetries, as well as individual preference.  The two exceptions are New York City, which draws the most out-of-state students, and Southern Tier, which is home to Cornell University and attracts the largest number of international students. Both of these outliers are intuitive; New York is a global city, and Cornell is also an ivy league college with excellent name recognition.  Mohawk Valley happens to draw a large number of non-regional in-state students, but that's likely because its comprehensive college, SUNY Oneonta, straddles the border with the Southern Tier region (and is also just upstate from New York City). 

```{r students, message = F, results = "hide", echo = TRUE}

#load student origin data
origins <- read_csv("suny_student_origins.csv")

#build stacked bar plot
ggplot(origins, aes(x = Region, y = Value, 
               fill = Key))+
  geom_col(position = position_stack(reverse = FALSE))+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1),
        axis.text = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7),
        axis.ticks = element_blank(),
        plot.caption = element_text(size = 6)
          )+
  labs(title = "SUNY Student Origins by Region",
       y = "Percent of Total Students",
       caption = "Data: Rockefeller Institute"
       )

```

## Part II.  How important are SUNY graduates to each region's workforce?

The relative importance of SUNY alumni - and regional schools - to sub-state labor markets varies widely.  For example, SUNY graduates make up a disproportionate number of total graduates in rural parts of the state like Mohawk Valley and North Country.  Graduates of regional SUNY schools make up over 50% of the North Country's total pool of residents with a degree, but total SUNY alums comprise more than 70% of area graduates.

Despite making up a smaller share of total graduates in the Capital District, SUNY graduates are about 35% of the total workforce there (which is highly educated on the whole).

Consistent with Florida's (2002) "Creative Class" thesis, SUNY graduates in New York City were mostly educated outside of the region, and the city succeeds in "pulling" college graduates into town. The opposite is true of Western New York; of the more than 60% of area graduates that are SUNY alums, about 55% were educated at regional schools.  This may be because SUNY has monopolized higher education in the region (though that seems unlikely as there are several other private colleges in and around Buffalo), or because a high proportion of graduates decide to remain in the area after graduation, or because the region struggles to attract college graduates from elsewhere (again, the challenge Florida describes in The Creative Class).

```{r suny alums grads, message = F, results = "hide", echo = TRUE}

#load grad retention data
grad_retention <- read_csv("grad_retention_percent_ttl_grads.csv")

#plot grad retention total graduates
ggplot(grad_retention, aes(x = Region, y = Value,
                           fill = Key))+
  geom_col(position = position_stack(reverse = FALSE))+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1),
        axis.text = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7),
        axis.ticks = element_blank(),
        plot.caption = element_text(size = 6),
        legend.key.size = unit(0.4, "cm")
  )+
  labs(title = "SUNY Alums as a Percent of Graduates in Region",
       y = "Percent of Total Graduates",
       caption = "Data: Rockefeller Institute")

```

``` {r suny alums workforce, message = F, results = "hide", echo = TRUE}

#load workforce grad retention data
grad_retention <- read_csv("grad_retention_percent_ttl_workforce.csv")

#build stacked bar plot
ggplot(grad_retention, aes(x = Region, y = Value,
                           fill = Key))+
  geom_col(position = position_stack(reverse = FALSE))+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1),
        axis.text = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7),
        axis.ticks = element_blank(),
        plot.caption = element_text(size = 6),
        legend.key.size = unit(0.4, "cm")
  )+
  labs(title = "SUNY Alums as a Percent of Regional Workforce",
       y = "Percent of Total Workforce",
       caption = "Data: Rockefeller Center")


```

## Part III.  Matching SUNY Graduates' Skills to Local Labor Markets

An important dimension of any public university's impact is whether the skills of their graduates match demand.  Here, I compare degrees held by alums living in the region as a percent of total regional jobs requiring a related degree to SUNY alums as a percent of total regional graduates (the dashed reference line).  This data is from the Rockefeller Institute, which compiled graduate major information from the International Postsecondary Education Data System (IPEDS) and job demand estimates from the NY Department of Labor.  
The series of bar plots below depict this skills (mis)match.


```{r skillsmatch, message = F, results = "hide", echo = TRUE}

#load Skills Match Data from CSV file
skills <- read_csv("suny_skills_match_by_region.csv")

#build faceted bar graphs w ref lines
ggplot(skills) + 
  geom_col(aes(x = Field , y = Degrees, fill = Field), 
           position = "dodge")+
  scale_fill_brewer(palette = "Paired")+
  labs(x = "Field", 
       y = "SUNY Degrees as a % of Jobs Requiring One",
       caption = "--- SUNY graduates as a proportion of total regional graduates")+
  theme_bw()+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
        )+
  geom_hline(aes(yintercept = Percent), 
             col = "Black", linetype = "dashed")+
  facet_wrap(~Region)+
  theme(legend.position = "bottom", 
        legend.box = "horizontal", 
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        legend.title = element_blank(),
        plot.caption = element_text(size = 7),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10)
        )+
  ggtitle("Matching SUNY Graduates' Skills to Area Labor Markets")

```

Overall, there is impressive consistency between SUNY graduates' skills and regional labor market demand.  But there are some notable mismatches.  In all cases, there is an oversupply of agriculture and art degrees.  The largest and most consistent undersupply is in hospitality.  Mohawk Valley, Long Island, and Western New York appear to have the least perfect relationship between skills supplied and skills demanded.  Central New York, Southern Tier, New York City, and the Capital District all fare better, though in different ways.

## Part IV.  Investigating the Skills Gap

Are these skills gaps primarily caused by broader market trends, such as higher demand for certain kinds of skills outside of the region?  Graduate retention and wage rates are a good proxy for the level of market power and degree of specialization alums hold, and help to illuminate whether students may have better job options elsewhere.        

```{r institutional level wages and retention, message = F, results = "hide", echo = TRUE}

#load wage data
wages <- read_csv("SUNY_locations_enrollment_wages.csv")

ggplot(wages, aes(x = Wages, y = Residents, col = Level))+
  geom_point()+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()+
  theme(legend.key.size = unit(0.4, "cm"))+
  ggtitle("Graduate Wages and New York Residency after 10 Years")+
  labs(x = "Wages 10 Years After Graduating", y = "% Graduates Living in New York After 10 Years",
       caption = "Data: SUNY Office of Institutional Research and Data Analytics")+
  theme(plot.caption = element_text(size = 7))

```

From the above scatterplot, it does appear as though graduates of Community, Comprehensive, and Technical schools are all more likely to remain in NY state after graduating. They also tend to make less money.  

However, while linear regressions reveal a significant and negative relationship between wages and percent of graduates who remain in New York State after graduating (the higher the grad's wages, the less likely they are to stay and work in NY), the overall explanatory power of this model is extremely weak (R2 = 0.03).


```{r Wages Residency Status Regression, message = F, results = "hide", warning = F, echo = TRUE}

programs <- read_csv("schools_programs_wages_residency.csv")

#filter schools that don't have programs or have fewer than 10 graduates
programs1 <- programs %>%
  filter(Wages != 0)
#linear regression wages residency status
residentwages <- lm(Residents~Wages, data = programs1)
#print summary
summary(residentwages)

```

```{r Wage Premiums by School and Field of Study, results = "hide", warning = F, message = F, echo = TRUE}

programs <- read_csv("schools_programs_wages_residency.csv")

#calculate wage premiums by school and field of study
programs_new <- programs %>%
  filter(Wages != 0) %>%
  mutate(wage_premium1 = Wages - Regional_Wages)%>%
  mutate(wage_premium2 = wage_premium1/Regional_Wages)

#linear regression wage premiums and NY residency
residentwagepremium <- lm(Residents~wage_premium2, data = programs_new)

#print summary
summary(residentwagepremium)

```

The results of these regressions suggest the skills gaps depicted above may also be a supply side issue.  After all, not all SUNY schools offer degree programs in all fields, or they may be missing one or more degree levels in a given region.   To dig into this a bit, I look at field of study, NY state residency, and wage rates in a few regions of interest: Western New York, Mohawk Valley, and Long Island.  

In Western New York, SUNY is undersupplying graduates in business and finance, computers and math, and hospitality.  One issue appears to be weak retention rates for graduates of the University at Buffalo (the only doctoral university in the area) in computers and math.  And no SUNY school in the area offers degrees in hospitality.

```{r Case Study Program Wages Western NY, message = F, results = "hide", echo = TRUE}

programs <- read_csv("schools_programs_wages_residency.csv")

#calculate average wages by program of study in each region
programs2 <- programs_new %>%
  group_by(Region) %>%
filter(Region == "Western New York")
  
#organize graphics
ggplot(programs2, aes(x = wage_premium2, y = Residents, col = Level))+
  geom_point()+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()+
  theme(legend.key.size = unit(0.4, "cm"))+
  ggtitle("Western New York: Graduate Wages and New York Residency after 10 Years")+
  labs(x = "Wage Premium 10 Years After Graduating", y = "% Graduates Living in New York After 10 Years")+
   theme_bw()+
  facet_wrap(~Program)+
  theme(legend.position = "bottom", 
        legend.box = "horizontal", 
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        legend.title = element_blank(),
        plot.caption = element_text(size = 7),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10)
        )

```

Mohawk Valley, a rural area in the center of the state, presents skills gaps in business and finance, education, health, hospitality, law and public service, and social services.

There are no doctoral universities in the region, so the gap in law and public service may be partially due to the absence of a law school (the only SUNY law school is at UB in Western New York).  A similar challenge is likely happening in the local healthcare industry.  Despite several degree programs at community, comprehensive, and technical schools, the absence of a medical school means demand for certain high-level skills is outstripping supply.

Retention rates are quite low for education and social service degree recipients, which may be driving the gap there.

```{r Case Study Mohawk Valley Mohawk Valley, message = F, results = "hide", echo = TRUE}
#calculate average wages by program of study in Mohawk Valley region
programs3 <- programs_new %>%
  group_by(Region) %>%
filter(Region == "Mohawk Valley")
  
#organize graphics
ggplot(programs3, aes(x = wage_premium2, y = Residents, col = Level))+
  geom_point()+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()+
  theme(legend.key.size = unit(0.4, "cm"))+
  ggtitle("Mohawk Valley: Graduate Wages and New York Residency after 10 Years")+
  labs(x = "Wage Premium 10 Years After Graduating", y = "% Graduates Living in New York After 10 Years",
       caption = "Data: SUNY Office of Institutional Research and Data Analytics")+
   theme_bw()+
  facet_wrap(~Program)+
  theme(legend.position = "bottom", 
        legend.box = "horizontal", 
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        legend.title = element_blank(),
        plot.caption = element_text(size = 7),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10)
        )

```

Long Island also presents skills gaps in business and finance, computers and math, education, health, hospitality, and technical trades.  Its doctoral institution, Stony Brook, is well known for its economics, math and computer science programs, and graduates may make more money elsewhere.  

```{r Case Study Program Wages Capital, message = F, results = "hide", echo = TRUE}

programs <- read_csv("schools_programs_wages_residency.csv")

#calculate average wages by program of study in each region
programs4 <- programs_new %>%
  group_by(Region) %>%
filter(Region == "Long Island")
  
#organize graphics
ggplot(programs4, aes(x = wage_premium2, y = Residents, col = Level))+
  geom_point()+
  scale_fill_brewer(palette = "Paired")+
  theme_bw()+
  theme(legend.key.size = unit(0.4, "cm"))+
  ggtitle("Long Island: Graduate Wages and New York Residency after 10 Years")+
  labs(x = "Wage Premium 10 Years After Graduating", y = "% Graduates Living in New York After 10 Years",
       caption = "Data: SUNY Office of Institutional Research and Data Analytics")+
   theme_bw()+
  facet_wrap(~Program)+
  theme(legend.position = "bottom", 
        legend.box = "horizontal", 
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        legend.title = element_blank(),
        plot.caption = element_text(size = 7),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10)
        )

```


## Conclusions and Additional Research Questions

In many ways, the SUNY state system is a model for human capital formation. Over the last 15 years, New York State's median educational attainment and have consistently exceeded national rates.  SUNY, which grants approximately 30% of postgraduate degrees in New York, may be largely to thank for this trend.

``` {r NY educational attainment, message = F, results = "hide", echo = TRUE}

regional_ed <- read_csv("suny_regional_education_tidy.csv")

#isolate NY and USA

NY <- regional_ed %>%
  filter(Region == "NY State")

USA <- regional_ed %>%
  filter(Region == "USA")

regional_ed1 <- regional_ed %>%
  filter(Region != "NY State") %>%
  filter(Region != "USA")

#plot education rates in each region

plot <- ggplot(data = regional_ed1, 
       aes(x = Year, 
           y = bach_or_higher, 
           col = Region))+ geom_line()+
  scale_color_brewer(palette = "Paired")

plot + geom_line(data = NY, 
                 aes(x = Year, y = bach_or_higher), 
                 col = "Black", size = 1)+
  geom_line(data = USA, 
            aes(x = Year, y = bach_or_higher), 
            col = "Black", linetype = "dashed", size = 1)+
  theme_bw()+
  theme(panel.grid.major = element_blank())+
  labs(x = "Year", 
       y = "% Population with Bachelor's or Higher", 
       title = "Educational Attainment in New York State",
       caption = "___ New York; -- US"
         )

```

Yet there is significant variation between regions, and SUNY's role in helping to attract, form, and redistribute human capital also varies widely across the state.  While the system has a relatively small effect on New York City, it plays a fundamental role in the economy of rural (Mohawk Valley, North Country) and post-industrial areas (Western and Central New York). 

While this study generated a few preliminary answers to the questions raised throughout, more work is needed to understand the relationship between specific SUNY institutions and regional labor markets.  A more precise measure of skills supplied/skills demanded is in order, something I intend to continue to refine (thanks to some great ideas from my GEO 511 classmates!) in the coming weeks and months.







