---
title: "Human Capital Formation and Redistribution as a function of Spatial Skills Mismatch "
author: Moira O'Neill
subtitle: Subtitle here if desired
---

# Introduction

In the United States, public colleges and universities play a critical role in attracting, concentrating, and redistributing human capital. 

Students often relocate from a dispersed set of hometowns to a few select localities to attend university. After graduating, many students will relocate to take up new jobs elsewhere. But some students who move to a new area for college will also stay in that community after they complete their degree (Winters 2011). This can have a compounding effect on local human capital stocks, and lead to productivity and wage gains for the entire workforce (Moretti 2004).

In rural and post-industrial communities with institutions of higher education but relatively low levels of human capital, graduate retention rates are likely to have a significant impact on a region’s overall competitiveness (Whisler et al. 2008). However, owing to the difficulty of measuring these effects and lack of available data, graduate retention has not been well-studied and remains poorly understood.

By comparing the majors of local college graduates to regional employment needs, this study will begin to address this gap by looking at college graduate’s decisions to stay or leave as a function of the following two-way effect:  first, between the kinds of human capital regional institutions are supplying, and second, the ability of a regional labor market to absorb and integrate supplied labor.


# Materials and methods

Narrative and most code will go here.  Describe what you are doing and show how to do it (with code).

You can do bullets like this:

* The first most important thing
* The second most important thing
* The third most important thing

You can do numbers like this:

1. The first most important thing
2. The second most important thing
3. The third most important thing

See [http://rmarkdown.rstudio.com/](http://rmarkdown.rstudio.com/) for all the amazing things you can do.


Here's my first code chunk.
```{r}
1+2
```

Load any required packages in a code chunk (you may need to install some packages):

```{r, message=F, warning=F}
library(tidyverse)
library(leaflet)
library(kableExtra)
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

## Download and clean all required data
```{r}
n=20
data=data.frame(x=runif(n,-180,180),
                y=runif(n,-60,60),
                size = runif(n, 5, 20),
                category = factor(
                  sample(letters[1:5], n, replace = TRUE)
                  ),
                value = rnorm(n))
```

```{r, results='asis'}
data %>% 
  slice(1:10) %>% #show only 1:n rows
  kable(digits=2,align="c")%>% #make table and round to two digits
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive")) #apply other formatting
```

Add any additional processing steps here.

# Results

Show tables, plots, etc. and describe them.

```{r, fig.width=6, fig.height=3, fig.cap="Map of completely random data"}
m <- leaflet(data) %>% 
  addTiles() %>% 
  addCircleMarkers(~x, ~y, radius = ~size,color = ~as.factor(category)) %>% 
  addPopups(~x[2], ~y[2], "Random popup")
m  # a map with the default OSM tile layer
```


```{r}
data %>% 
  ggplot(aes(x=x,y=y,col=category))+
  geom_point()
```

# Conclusions

What have you learned?  Are there any broader implications?

# References

Moretti, E. (2004). Estimating the social return to higher education: Evidence from longitudinal and repeated cross-section data. Journal of Econometrics 121 (1-2): 175-212.

Whisler, R.L. and B.S. Waldorf, G.F. Mulligan, D.A. Plane (2008). Quality of Life and the migration of the college educated: A life-course approach. Growth and Change 39 (1): 58-94.

Winters, John V. (2011). Why are smart cities growing? Who leaves and who stays. Journal of Regional Science 51 (2): 253-270.